---
title: "plsc30600_w2_ggn"
author: "your name here"
date: "2026-01-16"
output: html_document
---

Reference: Green, Donald P., Alan S. Gerber, and David W. Nickerson. " Getting out the
vote in local elections: Results from six door-to-door canvassing experiments."
The Journal of Politics 65.4 (2003): 1083-1096.
doi: https://doi.org/10.1111/1468-2508.t01-1-00126

Download replication data and data dictionary here: https://isps.yale.edu/research/data/d017

# Data Prep

```{r}
library(dplyr)

dat <- read.csv("/Users/maggiewang/Desktop/causal TA/sections/w2/green-et-al_2003.csv")
dat <- dat |> 
  filter(city == "Raleigh")
```

```{r}
# Check that the treatment and control are unequal
table(dat$treatmen) # yes, it's missing a 't'

# Note that the differences are due to different treatment assignments across "turfs"
prop.table(table(dat$turf, dat$treatmen), margin = 1)
```

```{r}
# recode to Ys and Ds
dat <- dat |> 
  mutate(
    y = 1*(voted01=="Voted"),
    d = 1*(treatmen == "Treatment")
  )
```

```{r}
# recode age as a discrete variable
dat <- dat %>%
  mutate(
    age_bin = cut(
      age,
      breaks = c(18, 30, 45, 65, 90),
      right = FALSE,
      labels = c("18–29", "30–44", "45–64", "65+")
    )
  )
```


# Exercises

1. Naive estimation: Estimate the ATE using the difference in sample means between the treated and control units


2. Estimate the ATE using post-stratification on race, party, sex, age, and turf


3. Estimation with the "true" propensity score

3.1 Generate the "true" propsensity score assuming that treatment is randomly assigned across turfs

3.2 Estimate the ATE with the "true" propensity score


4. Compare the different estimates. What do you find?



